---
title: "Overview of DelayedMatrixStats"
author: "Peter Hickey"
date: "Modified: 04 Oct 2017. Compiled: `r format(Sys.Date(), '%d %b %Y')`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE, setup}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", collapse = TRUE,
                      message = FALSE)
```

# Overview

`r Biocpkg("DelayedMatrixStats")` ports the `r CRANpkg("matrixStats")` API 
to work with *DelayedMatrix* objects from the `r Biocpkg("DelayedArray")` 
package. It provides high-performing functions operating on rows and columns of 
*DelayedMatrix* objects, including all subclasses such as *RleArray* (from the 
`r Biocpkg("DelayedArray")` package) and *HDF5Array* (from the 
`r Biocpkg("HDF5Array")`) as well as supporting all types of *seeds*, such as 
*matrix* (from the *base* package) and *Matrix* (from the `r CRANpkg("Matrix")` 
package).

# How can DelayedMatrixStats help me?

The `r Biocpkg("DelayedArray")` package allows developers to store array-like 
data using in-memory or on-disk representations (e.g., in HDF5 files) and 
provides a common and familiar array-like interace for interacting with these 
data.

The `r Biocpkg("DelayedMatrixStats")` package is designed to make life easier 
for Bioconductor developers wanting to use `r Biocpkg("DelayedArray")` by 
providing a rich set of column-wise and row-wise summary functions. Key to this 
are:

1. A common and familiar API
2. 'Seed-aware' optimized methods

We briefly demonstrate and explain these two features using a simple example.

## Example

We'll simulate some (unrealistic) RNA-seq read counts data from 10,000 genes 
and 20 samples and store it on disk as a *HDF5Array*:

```{r data_sim, message = FALSE}
library(DelayedArray)

x <- do.call(cbind, lapply(1:20, function(j) {
  rpois(n = 10000, lambda = sample(20:40, 10000, replace = TRUE))
}))
colnames(x) <- paste0("S", 1:20)
x <- realize(x, "HDF5Array")
x
```

Suppose you wish to compute the standard deviation of the read counts for each 
gene. 

You might think to use `apply()` like in the following:

```{r apply}
system.time(row_sds <- apply(x, 1, sd))
head(row_sds)
```

This works, but takes quite a while.

Or perhaps you already know that the `r CRANpkg("matrixStats")` package 
provides a `rowSds()` function:

```{r matrixStats, error = TRUE}
matrixStats::rowSds(x)
```

Unfortunately (and perhaps unsurprisingly) this doesn't work. 
`r CRANpkg("matrixStats")` is designed for use on in-memory *matrix* ojects. 
Well, why don't we just first realize our data in-memory and then use 
`r CRANpkg("matrixStats")`

```{r realization}
system.time(row_sds <- matrixStats::rowSds(as.matrix(x)))
head(row_sds)
```

This works and is many times faster than the `apply()`-based approach! However, 
it rather defeats the purpose of using a *HDF5Array* for storing the 
data since we have to bring all the data into memory at once to compute the 
result. 

Instead, we can use `DelayedMatrixStats::rowSds()`, which has the speed 
benefits of `matrixStats::rowSds()`[^speed] but without having to load the 
entire data into memory at once[^block_size]:

[^speed]: In fact, it currently uses `matrixStats::rowSds()` under the hood.
[^block_size]: In this case, it loads blocks of data row-by-row. The amount of 
data loaded into memory at any one time is controlled by the 
`DelayedArray.block.size` option; see `?DelayedArray::DelayedArray` for 
details. Notably, if the data are small enough (and the value of this option is 
largeenough) then all the data is loaded as a single block, but this approach 
generalizes and still works when the data are too large to be loaded into 
memory in one block.

```{r DelayedMatrixStats}
library(DelayedMatrixStats)

system.time(row_sds <- rowSds(x))
head(row_sds)
```

# Supported methods

The initial release of `r Biocpkg("DelayedMatrixStats")` supports the complete 
set of column-wise and row-wise `r CRANpkg("matrixStats")` API[^api].

[^api]: **NOTE:** `colWeightedSds()`, `colWeightedVars()`, `rowWeightedSds()`, and 
`rowWeightedVars()` will be implemented/reinstated after resolving 
[a bug in `matrixStats::weightedVar()`](https://github.com/HenrikBengtsson/matrixStats/issues/105)

```{r API, echo = FALSE}
matrixStats <- sort(grep("^(col|row)", 
                         getNamespaceExports("matrixStats"), 
                         value = TRUE))
# TODO: A better way of parsing DelayedMatrixStats NAMESPACE to get exported 
#       functions (will also need internal functions for api_df)
DelayedMatrixStats <- sort(grep("^(col|row)", 
                         getNamespaceExports("matrixStats"), 
                         value = TRUE))

api_df <- data.frame(
  Method = paste0("`", matrixStats, "()`"),
  `Block processing` = ifelse(matrixStats %in% DelayedMatrixStats,
                              "✔",
                              "❌"),
  `_base::matrix_ optimized` = 
    ifelse(sapply(matrixStats, existsMethod, signature = "matrix"), 
           "✔", 
           "❌"),
  `_Matrix::Matrix_ optimized` = 
    ifelse(sapply(matrixStats, existsMethod, signature = "Matrix"), 
           "✔", 
           "❌"),
  `_DelayedArray::RleArray_ (_SolidRleArraySeed_) optimized` = 
    ifelse(sapply(matrixStats, existsMethod, signature = "SolidRleArraySeed"),
           "✔", 
           "❌"),
  `_DelayedArray::RleArray_  (_ChunkedRleArraySeed_) optimized` = 
    ifelse(sapply(matrixStats, existsMethod, signature = "ChunkedRleArraySeed"),
           "✔", 
           "❌"),
  `_HDF5Array::HDF5Matrix_ optimized` = 
    ifelse(sapply(matrixStats, existsMethod, signature = "HDF5ArraySeed"),
           "✔", 
           "❌"),
  `_base::data.frame_ optimized` = 
    ifelse(sapply(matrixStats, existsMethod, signature = "data.frame"),
           "✔", 
           "❌"),
  `_S4Vectors::DataFrame_ optimized` =
    ifelse(sapply(matrixStats, existsMethod, signature = "DataFrame"),
           "✔", 
           "❌"), 
  check.names = FALSE)
knitr::kable(api_df, row.names = FALSE)
```

Please see the `r CRANpkg("matrixStats")` vignette ([available online](https://cran.r-project.org/package=matrixStats/vignettes/matrixStats-methods.html)) 
for a summary these methods.


# Benchmarking

# Roadmap
